---
title: "SpaxNMF"
page-layout: full
---

```{r}
index_summary_path <- "data/ngc1512-benchmark-summary.csv"
index_component_path <- "data/ngc1512-benchmark-components.csv"
index_bench <- if (file.exists(index_summary_path)) {
  read.csv(index_summary_path, stringsAsFactors = FALSE)
} else {
  NULL
}
index_components <- if (file.exists(index_component_path)) {
  read.csv(index_component_path, stringsAsFactors = FALSE)
} else {
  NULL
}

index_snapshot <- NULL
if (!is.null(index_bench)) {
  vanilla_row <- index_bench[index_bench$model == "Vanilla NMF", , drop = FALSE]
  spatial_row <- index_bench[index_bench$model == "Spatial NMF", , drop = FALSE]
  if (nrow(vanilla_row) && nrow(spatial_row)) {
    index_snapshot <- sprintf(
      "On the saved NGC1512 benchmark, the mixed spatial+sparse fit reduces the vanilla NMF MSE from %.3g to %.3g while raising the designated compact-component top-1%% concentration from %.3f to %.3f.",
      vanilla_row$mse,
      spatial_row$mse,
      vanilla_row$sparse_top1pct,
      spatial_row$sparse_top1pct
    )
  }
}
```

::: {.hero-shell}
<div class="hero-grid">
<div>
<p class="eyebrow">Spatially aware NMF for integral-field spectroscopy</p>

# Recover component spectra and spatial structure from IFU cubes

<p class="hero-copy">
SpaxNMF fits non-negative component spectra together with component-weight
maps, combining smooth spectral regularization, Gaussian spatial coupling, and
optional sparse priors for compact sources.
</p>

<div class="hero-actions">
  <a class="hero-button hero-primary" href="getting-started.html">Get Started</a>
  <a class="hero-button hero-secondary" href="benchmark.html">See the Benchmark</a>
</div>
</div>

<div class="hero-card">
<h3>What the spatial model adds</h3>

<ul class="hero-list">
<li>Shared non-negative component spectra recovered from the full cube.</li>
<li>Gaussian spatial coupling on component-weight maps for extended structure.</li>
<li>Optional sparse priors for compact-source components such as GC-like signals.</li>
<li>Real-cube diagnostics against PCA and vanilla NMF under matched preprocessing.</li>
</ul>

<p class="hero-card-note">
The full objective, kernel weighting, and optimization details are documented
on the <a href="model.html">Model</a> page.
</p>
</div>
</div>
:::

<div class="metric-ribbon">
<div class="metric-card">
<strong>Data View</strong>
<span>Spaxels × wavelength</span>
<p>Built around matrix factorization after reshaping IFU cubes.</p>
</div>
<div class="metric-card">
<strong>Spatial Prior</strong>
<span>Gaussian kernel</span>
<p>Nearby spaxels are coupled with PSF-like distance weighting.</p>
</div>
<div class="metric-card">
<strong>Compact Sources</strong>
<span>Optional sparse maps</span>
<p>Selected components can be driven toward GC-like or point-like structure.</p>
</div>
</div>

<div class="feature-grid">
<div class="feature-card">
<h3>Interpretability first</h3>
<p>
The factorization is non-negative, so each spaxel is reconstructed as a
positive combination of shared component spectra and spatially varying
component weights.
</p>
</div>
<div class="feature-card">
<h3>Designed for real cubes</h3>
<p>
The package includes helpers for IFU reshaping, metadata preservation, and
comparative diagnostics against PCA and vanilla NMF on the same target.
</p>
</div>
<div class="feature-card">
<h3>Method development ready</h3>
<p>
The current implementation supports smooth spectra, weighted spatial priors,
and sparse compact-source components inside the same torch optimization loop.
</p>
</div>
</div>

## Core Factorization

<div class="two-up">
<div class="equation-panel">
<p class="section-kicker">Factorization</p>
<h3>Shared spectra, spatially varying weights</h3>

$$
X \approx AS
$$

<p>
<code>X</code> is the spaxel-by-wavelength matrix, <code>A</code> is the
component-weight matrix, and <code>S</code> contains the component spectra.
</p>
</div>

<div class="equation-panel">
<p class="section-kicker">Spatial term</p>
<h3>Weighted coherence between neighboring spaxels</h3>

$$
w_{ij} = \exp\left(-\frac{d_{ij}^2}{2\sigma^2}\right)
$$

<p>
This term discourages noisy salt-and-pepper maps by penalizing strong changes
between nearby component weights while allowing distant spaxels to decouple.
</p>
</div>
</div>

## Quick Workflow

```{mermaid}
flowchart LR
  A["IFU cube<br/>nx × ny × nλ"] --> B["cube_to_matrix()"]
  B --> C["spaxel × wavelength matrix X"]
  C --> D["spax_nmf()"]
  D --> E["component-weight maps A"]
  D --> F["component spectra S"]
  D --> G["reconstruction AS"]
  E --> H["component_map()"]
  F --> I["plot(type = 'spectra')"]
  G --> J["predict(type = 'cube') / fitted()"]
```

## Real-Cube Benchmark

<div class="two-up">
<div class="info-panel">
<h3>NGC1512 galaxy + compact sources + sky</h3>
<p>
The current benchmark uses <code>NGC1512_cube_mockGCs.fits</code>, a mixed IFU
cube containing extended galaxy structure, compact GC-like sources, and sky
contamination in the same field. The comparison applies the same outer-spaxel
sky subtraction and matched preprocessing before fitting PCA, vanilla NMF, and
the mixed spatial+sparse model.
</p>

```{r}
if (!is.null(index_snapshot)) {
  cat(sprintf("<p class='benchmark-inline'>%s</p>", index_snapshot))
} else {
  cat("<p class='benchmark-inline'>The repository includes a reproducible real-cube benchmark that compares reconstruction quality and compact-component concentration under the same preprocessing.</p>")
}
```

<p class="hero-card-note">
Use the benchmark page for the full figure, saved metrics, and the exact run
configuration used to generate them.
</p>

<p><a href="benchmark.html">Open the real-cube benchmark</a></p>
</div>

<div class="info-panel">
<h3>What the benchmark is for</h3>
<p>
This benchmark is intended as a technical validation target rather than a
front-page showcase. It is useful for checking whether the mixed spatial+sparse
prior improves compact-component concentration and preserves acceptable
reconstruction quality under matched preprocessing.
</p>

<ul class="hero-list">
<li>PCA remains the rank-\(k\) squared-error baseline.</li>
<li>Vanilla NMF provides the non-negative comparison without spatial coupling.</li>
<li>SpaxNMF adds Gaussian spatial smoothing and optional sparse compact-source components.</li>
<li>The full figure, saved metrics, and run settings live on the benchmark page.</li>
</ul>

<p class="hero-card-note">
Use the benchmark page when you want the detailed diagnostic, not the main
landing-page overview.
</p>
</div>
</div>

## Main Entry Points

<div class="info-grid">
<div class="info-panel">
<h3>Fit models</h3>
<p><code>spax_nmf()</code> fits the torch model and returns spectra, weights,
reconstructions, and optimization history.</p>
</div>

<div class="info-panel">
<h3>Reshape data</h3>
<p><code>cube_to_matrix()</code> and <code>matrix_to_cube()</code> move between
IFU cubes and the matrix view required by the optimizer.</p>
</div>

<div class="info-panel">
<h3>Inspect results</h3>
<p><code>component_map()</code>, <code>component_spectrum()</code>,
<code>plot()</code>, and <code>plot_reconstruction()</code> provide quick
diagnostics.</p>
</div>

<div class="info-panel">
<h3>Benchmark on toy and real targets</h3>
<p>The website includes a controlled toy sanity check for standalone NMF and a
real mixed galaxy+GC+sky benchmark under matched preprocessing.</p>
</div>
</div>

## Installation

```r
# install.packages("remotes")
remotes::install_github("RafaelSdeSouza/SpaxNMF")
```

`torch` must be available in the local R installation.
