---
title: "IFU workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{IFU workflow}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette outlines the workflow that should anchor the package website and
any accompanying research note.

## 1. Start from a cube

The package assumes an `nx x ny x n_lambda` cube.

```{r}
library(SpaxNMF)

demo <- simulate_ifu_cube()
dim(demo$cube)
```

## 2. Reshape for fitting

```{r}
x <- cube_to_matrix(demo$cube)
dim(x)
```

## 3. Fit the unmixing model

```{r, eval = FALSE}
fit <- spectral_unmix(
  x,
  k = 3,
  lambda_smooth = 0.02,
  niter = 300,
  lr = 0.03
)
```

## 4. Inspect spectra and maps

```{r, eval = FALSE}
par(mfrow = c(1, 2))
plot(fit, type = "spectra", wavelength = demo$wavelength)
plot(fit, type = "maps", nx = demo$nx, ny = demo$ny)
plot_reconstruction(fit, x, n = 4, wavelength = demo$wavelength)
```

## 5. Extend to real data

For the next package demo layer, use one public MaNGA cube and document:

1. preprocessing choices
2. number of components
3. smoothness regularization
4. comparison baseline: PCA, vanilla NMF, and spatial NMF
5. astrophysical interpretation of each component

The synthetic workflow remains useful even after that, because it gives a
stable, fast, fully reproducible example for the package website.
